Supermartingales and Termination Analysis of Statistical PCF

### 2021-10-18 LICS21/antitone.tex
Start adding the new stronger (yet still partial) completeness result

The following theorem does not prove completeness, but is suggestive in that direction:
%\begin{restatable}{therm}{antitoneSeemsComplete}
\begin{therm}
\label{thm:towards completeness}
For any stopping time $T$ which is almost surely finite, if $(\calF_n)_n$ is the coarsest filtration to which $T$ is adapted, then there is a supermartingale $(Y_n)_n$ adapted to $(\calF_n)_n$ and an antitone function $\epsilon$ such that $(Y_n)_n$ is an antitone ranking supermartingale with respect to $T$ and $\epsilon$.
\end{therm}
%\end{restatable}

\begin{proof}

If $T$ is bounded by $b$ a.s.~(for $n$ constant), the statement is trivially true by taking $\epsilon$ to be constantly 1, and $Y_n = b - n$.
Otherwise, let $(t_n)_{n \geq 0}$ be defined recursively such that 
\begin{align*}
t_0 &= 0\\
t_{n+1} &> t_n\\
\mathbb P[T > t_{n+1}] &\leq \frac 1 4 \mathbb P[T > t_n].
\end{align*}
We will then define a supermartingale $(Y_n)_{n \geq 0}$ and nonrandom sequence $(y_n)_{n \geq 0}$ such that $Y_n = 0$ iff $T < n$, $Y_n = y_n$ iff $T \geq n$, and $y_n \geq 2^k$ for $n \geq t_k$.

The assumption that $(\mathcal F_n)_{n \geq 0}$ is the coarsest filtration to which $T$ is adapted is used in the equality between the third and fourth lines here. This condition is the main reason that this proof does not extend directly to a completeness result for antitone ranking functions.

>>>>

The following theorem does not prove completeness, but is suggestive in that direction:
\begin{therm}
\label{thm:towards completeness supermartingales}
For any stopping time $T$ which is almost surely finite, adapted to a filtration $(\calF_n)_n$, then there is a supermartingale $(Y_n)_n$ adapted to $(\calF_n)_n$ and an antitone function $\epsilon$ such that $(Y_n)_n$ is an antitone ranking supermartingale with respect to $T$ and $\epsilon$.
\end{therm}

\begin{proof}
First, take the case where $T$ is bounded by some integer $b$. Setting $\epsilon(x) = 1$ and $Y_n = \max(b-n, 0)$ trivially gives an antitone ranking supermartingale for $T$.

Assume, for the other case, that $T$ is unbounded. Let $f(n) = \mathbb P[T < n]$, let $g : [0,1) \to \mathbb N$ be a non-decreasing function such that $g(f(n)) \geq n$ \akr{This was originally $f(g(x)) \geq x$. Maybe that was right? Maybe I need both. I think that would be fine. I'll see how the rest of the proof goes.} (essentially just taking the inverse of $f$ and filling in the gaps) and let $\epsilon : \mathbb R_{\geq 0} \to \mathbb R_{>0}$ be a convex antitone function less than or equal to $(g(1-2^{-x}))^{-1}$. Then we can define the supermartingale $(Y_n)_n$ by $Y_n = \mathbb E[-\log_2(1-f(T-n)) \mid \calF_n]$. Because $T$ is unbounded, $f$ never reaches 1 therefore the logarithm is finite.

In order to show that the expectation $\mathbb E[-\log_2(1-f(T-n))]$ converges, consider its definition as a Lebesgue integral (using the fact that the argument of the expectation here is non-negative):
\begin{align*}
& \mathbb E[-\log_2(1-f(T-n))] \\
= & \int_0^\infty \mathbb P[-\log_2(1-f(T-n)) > x] \, \mathrm d x \\
= & \int_0^\infty \mathbb P[f(T-n) > 1-2^{-x}] \, \mathrm d x \\
\leq & \int_0^\infty \mathbb P[f(T) > 1-2^{-x}] \, \mathrm d x \\
= & \int_0^\infty \mu(\{a \mid \mu(\{b \mid T(b) < T(a)\}) > 1-2^{-x}\}) \, \mathrm d x \\
\end{align*}
Call the set $\{a \mid \mu(\{b \mid T(b) < T(a)\}) > 1-y\}$ $A_y$. Because the predicate $\mu(\{b \mid T(b) < T(a)\}) > 1-y$ is monotonic in $T(a)$, all elements of $A_y$ have higher values of $T$ than all non-elements of $A_y$. Let $a$ be the element of $A_y$ with the minimum value of $T$. Then $1-y < \mu(\{b \mid T(b) < T(a)\}) = \mu(A_y^\complement) = 1 - \mu(A_y)$ therefore $\mu(A_y) < y$. The integral is therefore less than
\begin{align*}
\int_0^\infty 2^{-x} \, \mathrm d x < \infty
\end{align*}
therefore $(Y_n)_n$ is well defined.

The sequence is clearly non-negative and adapted to the filtration, so the only condition left to check is the progress condition.
\begin{align*}
& \epsilon(Y_n) \\
= & \epsilon(\mathbb E[-\log_2(1-f(T-n)) \mid \calF_n]) \\
\leq & \mathbb E[\epsilon(-\log_2(1-f(T-n))) \mid \calF_n] \\
\leq & \mathbb E[(g(1-2^{\log_2(1-f(T-n))}))^{-1} \mid \calF_n] \\
= & \mathbb E[(g(f(T-n)))^{-1} \mid \calF_n] \\
\end{align*}

\akr{I haven't finished writing the proof.}

\end{proof}

This construction can be applied to produce antitone ranking functions in most but not all cases. More than simply every reachable term being AST, they must all terminate uniformly quickly. The restriction, more specifically, is as follows:

\begin{therm}
\label{thm:towards completeness}
For any closed term $M$, if there is some function $t : \mathbb N \to [0,1]$ that tends to 1 such that for any $N \in \mathit{Rch}(M)$, the probability that $N$ terminates within $n$ steps is at least $t(n)$, then $M$ is antitone rankable.
\end{therm}

\begin{proof}
\end{proof}

This theorem is not able to account for all antitone rankable terms.

\begin{example}
Let $M = (\tY \lambda f s. \tif{-\tsample < 0}{(\tY \lambda r n. \tif{n < 0}{\underline 0}{r (n - 1)}) s \lfloor \frac 1 \tsample \rfloor}{f (\lambda n. 2^{s n})}) (\lambda n. n)$. It first draws samples repeatedly until one of them is not exactly 0, then choses a random number, applies a function to it, and then loops that many times. Every time the sample drawn is exactly 0, the function gets exponentially faster growing. It therefore doesn't satisfy the conditions of Theorem \ref{thm:towards completeness} because the term reached after $n$ draws of 0 terminates slower as $n$ increases. However, it is antitone rankable nevertheless.
\end{example}
